Type-Safe Modular Parsing

### 2016-12-27 paper/sections/Overview.tex
Overview ready for a second review.

\paragraph{Extensibility} Similarly, we extend the grammar by adding lambda abstractions as a new case. The new parser also extends the old one by adding \inlinecode{pLam} on it. Notice that the old code is not touched, hence separate compilation is obtained.

\lstinputlisting[linerange=82-88]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=OVERVIEW_OPEN_EXT

@@

All recursive calls of parsing `an expression' are called via the explicit parameter \inlinecode{self}. Once the \inlinecode{fix} function is applied, these calls will be updated appropriately. Parsing functions such as \inlinecode{pApp} do not need to be rewritten. As a consequence, this parser recognizes all valid expressions of the new grammar, including \inlinecode{(\\x.x)} \inlinecode{(\\y.y)} which cannot be parsed in the previous section.

\lstinputlisting[linerange=94-94]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=OVERVIEW_OPEN_EXT_USE

@@

Although delegation enables OO extensibility, the use of an OO class
hierarchy makes the addition of new operations over the AST
problematic. Object Algebras~\cite{Oliveira2012} enable us to solve this problem, and
offer high flexibility in the choice of operations to be performed
over the AST. It also makes parsing with multiple sorts of syntax easier.

\paragraph{Parsing with Object Algebras} Using Object Algebras, the abstract syntax of the language of variables and applications is defined as below.

@@

Then we are able to define operations over the syntax in an extensible way. Here we write a pretty printing operation for this language.

@@

And also parser for it as below. Notice the parsing function \inlinecode{pExpr} takes an argument of type \inlinecode{ExprAlg[E]}, that means it accepts any instance of the algebra interface \inlinecode{ExprAlg}.

\lstinputlisting[linerange=22-28]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_PARSER

\paragraph{Extensibility of Syntax} Following previous examples, the code below shows how to extend the language to support lambda abstractions.

\lstinputlisting[linerange=32-46]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_EXT

For easily combining parsers using Object Algebras as ASTs, we have a special combinator \huang{we need to rename our combinator} defined in our library.

To use the parser, we must provide an algebra instance as the operation to construct the parsing results. In the code below, we use the pretty printing operation \inlinecode{LambdaPrint} as the algebra instance, so that the parsing result is a pretty printed AST.

\lstinputlisting[linerange=55-60]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_USE

\paragraph{Extensibility of Operations} With Object Algebras, operations over ASTs can also be extended in a modular way. Here is an example of collecting free variables from an expression. We can use this operation during parsing and obtain a set of free variables.

\lstinputlisting[linerange=74-80]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_EXT_OP

@@

\begin{tabular}{m{0.4\linewidth}m{0.4\linewidth}}
\setlength{\grammarindent}{5em}
\begin{grammar}
<type> ::= `int' \alt <type> `->' <type>
\end{grammar}
&
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <ident> \alt <expr> <expr> \alt `\\' <ident> `.' <expr>
\end{grammar}
\end{tabular}

Using Object Algebras, we can easily distinguish them in an extensible way, just by adding an extra type parameter.

\lstinputlisting[linerange=65-69]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_MULTI_SYNTAX

>>>>

\paragraph{Extensibility} Similarly, we extend the grammar by adding variables as a new case. The new parser also extends the old one by adding \inlinecode{pVar} on it. Notice that the old code is not touched, hence separate compilation is obtained.

\lstinputlisting[linerange=82-85]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=OVERVIEW_OPEN_EXT

@@

All recursive calls of parsing `an expression' are realized via the explicit parameter \inlinecode{self}. Once the \inlinecode{fix} function is applied, these calls will be updated appropriately. Parsing functions such as \inlinecode{pAdd} do not need to be rewritten. As a consequence, this parser recognizes all valid expressions of the new grammar, including \inlinecode{"1 + x"} which cannot be parsed previously.

\lstinputlisting[linerange=91-91]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=OVERVIEW_OPEN_EXT_USE

@@

Although delegation enables OO extensibility, the use of an OO class
hierarchy makes the addition of new operations over the AST
problematic. Modification on existing code breaks the modularity. This
refers to the famous \textit{Expression Problem} []. For challenge 2, suppose now we want to
support evaluation on expressions, in Scala one attempt would be extending \lstinline{Expr} with
evaluation:

\lstinputlisting[linerange=111-111]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=OVERVIEW_ATTEMPT_EXPRWITHEVAL
But then the programmer has to define new classes for literals and additions that extend old ones and \lstinline{EvalExpr}. Specifically,
the critical point is that existing parsing code only generates \lstinline{Expr} objects with pretty printing, so modification on parsing is again required.

In contrast, Object Algebras~\cite{Oliveira2012} enable us to solve this problem, as it separates data variants and
operations, to
offer high flexibility in the choice of operations to be performed
over the AST. It also makes parsing with multiple sorts of syntax easier.

\paragraph{Parsing with Object Algebras} Using Object Algebras, the abstract syntax of the expression language is defined as below.

@@

Then we are able to define operations over the syntax in a modular way. For instance, pretty printing operation can be realized as:

@@

And also parser for it as below. Notice the parsing function \inlinecode{pExpr} now takes an argument of type \inlinecode{ExprAlg[E]}, which means it accepts any instance of \inlinecode{ExprAlg}. Such an instance can for example be \lstinline{Print}, which takes charge of operations on expressions, and parsing code only does delegation.

\lstinputlisting[linerange=22-27]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_PARSER

\haoyuan{I'm trying to avoid using terminology like algebra and algebra interface.}

\paragraph{Extensibility of Syntax} Following previous examples, the code below shows how to extend the language with variables.

\lstinputlisting[linerange=31-44]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_EXT
Note that we use the new combinator \lstinline{<|>} defined in our library, to compose two parsing functions into one. The function \lstinline{pExtExpr} has the compound type \lstinline{ExprAlg[E] with VarAlg[E]} in its argument, where two language ASTs are combined to represent the new language.

To use the parser, we must provide an algebra instance as the operation to construct the parsing results. In the code below, we use the combination of \lstinline{Print} and \lstinline{PrintVar}, so that the parsing result is a pretty printing of the AST.

\lstinputlisting[linerange=53-58]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_USE

\paragraph{Extensibility of Operations} With Object Algebras, operations over ASTs can also be extended in a modular way. Here is an example of collecting free variables from an expression. We can feed this operation to the parser and obtain a set of free variables.

\lstinputlisting[linerange=72-78]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_EXT_OP

\haoyuan{I think in our library language ASTs are composed using intersection, not inheritance.}

@@

\begin{tabular}{m{0.45\linewidth}m{0.45\linewidth}}
\setlength{\grammarindent}{5em}
\begin{grammar}
<type> ::= `int' \alt <type> `->' <type>
\end{grammar}
&
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::=  `\\' <ident> `:' <type> `.' <expr>
\end{grammar}
\end{tabular}

Using Object Algebras, we can easily distinguish them just by adding an extra type parameter. The abstract syntax is
presented below:

\lstinputlisting[linerange=63-67]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_MULTI_SYNTAX
In the next sections, we will discuss them with more details.
